Dynastic Extinctions and Urban Economic Development: \\Evidence from the Holy Roman Empire\footnote{I thank Davide Cantoni, Matthias Weigand, and my RA colleagues at the Chair of Economic History for helpful comments and feedback.}

### 2023-06-25 paper/paper.tex
#30 almost finish feedback from Lara + Mama

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win with state power, and state power is simply free resources, i.e. resources above subsistence. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote{The basic idea of Darwinian selection has previously been borrowed by economists - for example \cite{galor2002} do the literal version, where humans are subject to selection and accumulate favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that the rich out-bred the poor and passed favourable traits on to their offspring (in Europe but not elsewhere for some reason.)}

@@

The Holy Roman Empire was a powerful monarchy that existed in Central Europe from the Middle Ages until its dissolution in 1806. Though ruled by the Holy Roman Emperor at the Imperial level, it consisted of many small and large territories with their own regional rulers. Before Germany first became a republic in 1918, it is not helpful to think of "states" in the modern sense: while territories may have existed as political entities on paper, they were ruled by aristocrats who handled power in their familial structures \citep[pp. 2-3]{pt2}. These aristocratic families defined the territorial history of the HRE: territories were split by inheritances, joined by marriages and extinctions, and driven to war by struggles for dominance among the noble lineages ruling them.

When I say "state" in the context of this paper I mean its early modern German equivalent --- the territory of a lineage of rulers, or the lineage itself.

@@

The record of construction events is unfortunately not exhaustive and mostly contains notable buildings --- churches, palaces, fortifications, malls, schools, libraries, opera houses, gardens, and the like. It contains residential buildings of some noble families and notable citizens but not of the general population. Even so, there are theoretical justifications for using construction activity as a proxy for growth: If buildings are a normal good, then demand for them should increase with income. While the types of buildings listed do seem to represent state and elite consumption more than general economic activity, the income of these groups is also dependent on the broader economy --- an upturn will increase tax revenues and by extension the disposable income of the state. On a basic level we would expect more new buildings, of any type, in cities that are booming (although the error term is likely large - we might also expect more new buildings in cities that were devastated by war or environmental catastrophes).

@@

Note that conflict \textit{incidents} are distinct from conflict as a driver of improvements in institutions. Conflict incidents are events recorded at the city level that often involve the destruction of life or property in that city. The statement "conflict causes weak states to disappear "

@@

The raw \textit{Territorial Histories} dataset starts out with 1,471,934 yearly observations. I lose two observations per city --- the first and the last year --- while coding switches, for a total of 4,780. I aggregate the remaining 1,467,154 yearly observations into 30,980 observations with 50-year spacing. Of these I drop 16,110 that are not covered by the construction data. Finally, the baseline versions of all specifications in this paper restrict the sample to cities with no more than two lifetime switches. This will remove another 9,975 observations, leaving 6,635 in the baseline sample.

@@

This is a standard TWFE regression with staggered adoption and dynamic effects. Treatment events are \textit{lasting} switches to another state, i.e. a switching event only counts as treatment if the city belongs to a different state at the beginning of the next period.\footnote
{
    If a city undergoes multiple lasting switches, I create duplicate entries for each different $e^{NewState}_i$. I exclude cities that switch states (permanently or not) more than twice over the sample period. I report results obtained with different exclusion thresholds in Appendix BLANK.
}
This is because I want the coefficients $\pi_\tau$ to capture the effects of the new state's governance, not just of the political and economic turmoil that may be associated with a transition of power. I separately control for other switches by including the term $\mathbf{1}(S_{it} > 0)$ on the right-hand side so that the coefficient $\beta$ absorbs the effect of switching states in the same period.

Since the outcome is binary in this case, we can interpret Equation \eqref{eq:sw22} as a linear probability model: we model $construction_{it}$ as a random variable that follows a Bernoulli distribution with a probability parameter $p$ that is conditional on the right-hand side variables. This way we can interpret the values of coefficients as differences in the probability of construction in a given period. The coefficient $\pi_\tau$ for given $\tau$ is the difference in the probability of construction between two cities $i$ and $j$ with identical time-invariant characteristics, in the same period, who undergo the same kind of switching turmoil in this period. The only difference is that one city underwent a lasting switch to another state $\tau$ periods ago, while the other did not.

@@

In the best case, if the identifying assumption is satisfied, the estimates from the previous regression tell us what the effect of switching states is --- however, they cannot tell us why. In the following I attempt to identify the effect of state power, or more accurately, the effect of the difference in selection for state power between conquests and successions.

@@

For clarity I move from the event study to a basic difference-in-differences specification; this way there is only a single coefficient of interest. I simply replace the lead and lag dummies with the post-treatment indicator $\mathbf{1}(t \ge e^{NewState}_i)$. Another minor change is that this equation features a vector of city-year level control variables $\mathbf{X}_{it}$ --- however, in the baseline version of this specification I control only for the dummy $\mathbf{1}(S_{it} > 0)$ that indicates switches in the same period, exactly like in Equation \eqref{eq:sw22}. In a robustness check I will additionally control for conflict incidents.

@@

If this assumption holds true, the effect of state power should manifest itself in the difference between the trend of cities acquired by conquest and the trend of cities acquired by succession. Note that successions are the omitted category --- there is no interaction term between the post-treatment indicator and $\mathbf{1}(Succession_{it})$ in Equation \eqref{eq:baseline} --- so that the coefficient $\theta_{conquest}$ captures exactly what we want: the difference in trends after conquests vis-a-vis successions.

The econometric assumption necessary to allow causal interpretation of $\theta_{conquest}$ is a different version of the parallel trends assumption. In particular, it must hold that, had a conquered city i counterfactually been acquired by succession rather than conquest, it would have trended the same as the other cities acquired by succession. BLANK.

@@

The regression in Column (1) used all construction activity as the outcome, while Columns (2), (3), and (4) use subsets as defined in Section \ref{sec:data}. Almost all the coefficients in the last three columns are insignificant, suggesting that neither of these specific subsets of construction is driving the results.

@@

Comparing Column (2) to (1), and (4) to (3), we can see that it makes no difference whether we control for switches in the same period or not; the estimates are barely distinguishable. When we control for conflict, however, the coefficient $\theta_{conquest}$ suddenly turns negative, highly significant and large in magnitude. If we trust these regressions, the probability of observing construction in a different year increases by 15 percentage points \textit{less} after conquests relative to successions.

@@

Other explanations for no effect:
\begin{itemize}
    \item states within the HRE may be too similar / competition between them not large enough
    \item successions may also select for state power - maybe you get to enter more advantageous marriages if you are the leader of a powerful state
    \item cities are normally conquered one at a time, while successions usually see many cities switch at the same time - maybe there is a crowding-out of the positive effect, e.g. because the ruler likes one of his newly acquired cities most and neglects the others
    \item the data is just not that well suited to the task; within-R2 is super low in Table \ref{tab:baseline_50y}
    \item had I found an effect, it would have been difficult to show that

\end{itemize}

>>>>

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win with state power, and state power is simply free resources, i.e. resources above subsistence. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their favourable traits on to their offspring.
}

@@

The Holy Roman Empire was a powerful monarchy that existed in Central Europe from the Middle Ages until its dissolution in 1806. Though ruled by the Holy Roman Emperor at the Imperial level, it consisted of many small and large territories with their own regional rulers. Before Germany first became a republic in 1918, it is not helpful to think of "states" in the modern sense: while territories may have existed as political entities on paper, they were ruled by aristocrats who handled power in their familial structures \citep[pp. 2-3]{pt2}. These aristocratic families defined the territorial history of the HRE: territories were split by inheritances, joined by marriages and extinctions, and driven to war by struggles for dominance among the noble lineages ruling them. For this reason, when I say "state" in the context of this paper I mean its early modern German equivalent --- the territory of a lineage of rulers, or the lineage itself.

@@

The historical record of construction events is unfortunately not exhaustive and mostly contains notable buildings --- churches, palaces, fortifications, malls, schools, libraries, opera houses, gardens, and the like. It contains residential buildings of some noble families and notable citizens but not of the general population. Even so, there are theoretical justifications for using construction activity as a proxy for growth: If buildings are a normal good, then demand for them should increase with income. While the types of buildings listed do seem to represent state and elite consumption more than general economic activity, the income of these groups is also dependent on the broader economy --- an upturn will increase tax revenues and by extension the disposable income of the state. On a basic level we would expect more new buildings, of any type, in cities that are booming (although the error term is likely large - in the short run we might also expect more new buildings in cities that were destroyed by war or environmental catastrophes).

@@

Note that conflict \textit{incidents} are distinct from conflict as a driver of improvements in institutions. Conflict incidents are events recorded at the city level that often involve the destruction of life or property in that city. By contrast, in the hypothesis "conflict selects for states with higher-quality institutions" I mean conflict at the macro level, as a process that determines the fates of entire territories.

@@

The raw \textit{Territorial Histories} dataset starts out with 1,471,934 yearly observations. I lose two observations per city --- the first and the last year --- while coding switches, for a total of 4,780. I aggregate the remaining 1,467,154 yearly observations into 30,980 observations with 50-year spacing. Of these I drop 16,110 that are not covered by the construction data. Finally, following \cite{schoenholzer2022}, for the baseline versions of all specifications I restrict the sample to cities with no more than two lifetime switches.\footnote
{
I relax this restriction in a robustness check.
} 
This will remove another 9,975 observations, leaving 6,635 in the baseline sample.

@@

This is a standard TWFE regression with staggered adoption and dynamic effects. Treatment events are \textit{lasting} switches to another state, i.e. a switching event only counts as treatment if the city belongs to a different state at the beginning of the next period.\footnote
{
    If a city undergoes multiple lasting switches, I create duplicate entries for each different $e^{NewState}_i$.
}
This is because I want the coefficients $\pi_\tau$ to capture the effects of the new state's governance, not just of the political and economic turmoil that may be associated with a transition of power. I separately control for other switches by including the term $\mathbf{1}(S_{it} > 0)$ on the right-hand side so that the coefficient $\beta$ absorbs the effect of switching states in the same period.

Since the outcome is binary in this case, we can interpret Equation \eqref{eq:sw22} as a linear probability model: we model $construction_{it}$ as a random variable that follows a Bernoulli distribution with a probability parameter $p$ that is conditional on the right-hand side variables. This way we can interpret the values of coefficients as differences in the probability of construction in a given period. The coefficient $\pi_\tau$ for given $\tau$ is the difference in the probability of construction between two cities $i$ and $j$ with identical time-invariant characteristics, in the same period, who undergo the same kind of switching turmoil in this period. The only difference is that one city switched permanently to another state $\tau$ periods ago, while the other did not.

@@

In the best case, if the identifying assumption is satisfied, the estimates from the previous regression tell us what the effect of switching states is --- however, they cannot tell us why. In the following I attempt to estimate the effect of state power, or more accurately, the effect of the difference in selection for state power between conquests and successions, on construction activity.

@@

For clarity I have moved from the event study to a basic difference-in-differences specification; this way there is only a single coefficient of interest. I simply replace the lead and lag dummies with the post-treatment indicator $\mathbf{1}(t \ge e^{NewState}_i)$. Another minor change is that this equation features a vector of city-year level control variables $\mathbf{X}_{it}$ --- however, in the baseline version of this specification I control only for the dummy $\mathbf{1}(S_{it} > 0)$ that indicates switches in the same period, exactly like in Equation \eqref{eq:sw22}. In a robustness check I will additionally control for conflict incidents.

@@

If this assumption holds true, the effect of state power should manifest itself in the difference between the trend of cities acquired by conquest and the trend of cities acquired by succession. Note that successions are the omitted category --- there is no interaction term between the post-treatment indicator and $\mathbf{1}(Succession_{it})$ in Equation \eqref{eq:baseline} --- so that the coefficient $\theta_{conquest}$ captures exactly what I want to estimate: the difference in trends after conquests vis-a-vis successions.

The econometric assumption that must hold true for $\theta_{conquest}$ to be a causal effect is a different version of the parallel trends assumption. In particular, it must hold that, had a conquered city i counterfactually been acquired by succession rather than conquest, it would have trended the same as the other cities acquired by succession. BLANK.

@@

The regression in Column (1) uses all construction as the outcome, while Columns (2), (3), and (4) use the subsets of construction activity defined in Section \ref{sec:data}. Almost all the coefficients in the last three columns are insignificant, suggesting that neither of these specific subsets of construction is driving the overall results. Curiously, though, 50 years after switching there is a small, transient, but significant increase in construction of public goods. Since many of these would have been provided by the state, an explanation could be that rulers are initially reluctant to invest into their new territory --- the coefficient for the 50 years immediately following the switch ($\tau = 0$) is insignificant but negative --- but catch up once their rule is established.

@@

Comparing Column (2) to (1), and (4) to (3), we can see that it makes no difference whether we control for switches in the same period or not; the estimates are barely distinguishable. When we control for conflict, however, the coefficient $\theta_{conquest}$ suddenly turns negative, highly significant and large in magnitude. If we trust these regressions, the probability of observing construction in a given year increases by 15 percentage points \textit{less} after conquests relative to successions.

@@

Other explanations for no effect:
\begin{itemize}
    \item states within the HRE may be too similar / competition between them not large enough
    \item successions may also select for state power - maybe you get to enter more advantageous marriages if you are the leader of a powerful state
    \item cities are normally conquered one at a time, while successions usually see many cities switch at the same time - maybe there is a crowding-out of the positive effect, e.g. because the ruler likes one of his newly acquired cities most and neglects the others
    \item the data is just not that well suited to the task; within-R2 is super low in Table \ref{tab:baseline_50y}
\end{itemize}

Improvements / future research:
\begin{itemize}
    \item differences in timing - do early switchers in my sample experience different effects than late switchers?
    \item lag controls because construction takes a while and the years probably record completion
\end{itemize}
